\section{Conclusion}
\label{sec:conclusion}

We asked whether a population of $N$ self-training LLMs has a
\emph{minimum viable population} below which a shared synthetic-text
substrate inevitably collapses, and what counts as enough
differentiation between agents. Two complementary harnesses --- a
fine-tuning ecosystem of \distilgpt copies and a retrieval ecosystem
of frozen API LLMs --- run on a common metric suite give a unified
answer with two regimes.

\para{Bottom line.}
For a frozen-parameter ecosystem, $\mathbf{N{=}2}$ agents from
distinct pretraining families preserve diversity for at least twelve
iterations (slope of mean pairwise distance: $-0.002 \pm 0.004$,
$p = 0.44$). For a self-training ecosystem at fixed per-ecosystem
compute, even $\mathbf{N{=}16}$ homogeneous agents are insufficient to
halt collapse over six iterations; the perplexity slope falls
$\sim 5\times$ from $N{=}1$ to $N{=}16$ but never reaches zero. Across
both regimes, the only individuality axis that meaningfully reduces
collapse is the underlying pretraining model family: at fixed $N{=}4$
it preserves $1.9\times$ the inter-agent distance of prompt-, retrieval-,
or seed-based diversity.

\para{Key takeaway.}
Population diversity that defends against ecosystem collapse must come
from diversity in the prior, not the conditioning, and the
training-time link between agents is what determines whether collapse
can be halted at any finite $N$ at all.

\para{Future work.}
Four directions follow. (1) Push the fine-tuning sweep to $N{=}64$ or
$N{=}128$ to see whether the perplexity-vs-iteration slope is
asymptotically zero for some finite $N$, or whether it remains
positive in the limit. (2) Combine the fine-tuning ecosystem with
$N$ different pretrained checkpoints to test whether the architecture-
diversity benefit observed in the \rag harness transfers to the
parameter-updating regime. (3) Combine population diversity with
\citet{gerstgrasser2024accumulating}-style data accumulation. (4) Run
\rag horizons of $T{=}50$ or $T{=}100$ to check whether $N{=}2$
model-family ecosystems eventually decay.

\paragraph{Reproducibility.}
Per-iteration metrics, generated-text samples, and per-run
configurations are saved to JSONL logs. Hardware: 1$\times$NVIDIA
RTX A6000 for E1; CPU + OpenRouter API for E2/E3.

\paragraph{Broader impact.}
Our results refine, rather than overturn, existing concerns about
LLM-generated text contaminating future training corpora. Frozen-model
ecosystems with architectural diversity are robust; self-training
loops are not, even with substantial population diversity. Because
the result speaks to AI deployment policy, the most important caveat
is the limitation on model scale: \distilgpt is small, and we cannot
rule out qualitatively different dynamics at the scale of frontier
foundation models.
