\section{Methodology}
\label{sec:methodology}

\subsection{Problem formulation}
\label{sec:problem}

Let $\{A_1, \ldots, A_N\}$ be a population of $N$ generative agents
sharing a common text substrate $\gS_t$ at iteration $t$. At each step,
every agent $A_i$ produces a new text slice $\gS_t^{(i)}$ and the pool
is updated $\gS_{t+1} \leftarrow \mathrm{update}(\gS_t,
\{\gS_t^{(i)}\}_i)$. We say the ecosystem \emph{collapses on
metric $M$} over horizon $T$ if the sequence $M(\gS_1), M(\gS_2),
\ldots, M(\gS_T)$ has a regression slope significantly different from
zero in the worsening direction. The \emph{minimum viable population}
on $M$ at horizon $T$, $\mvp_M(T)$, is the smallest $N$ for which the
slope's null hypothesis of ``no decay'' cannot be rejected at
$p > 0.05$, or, for lower-better metrics, the smallest $N$ for which
the terminal value remains within
$\mathrm{horizon\_frac} \cdot (M(\gS_T)|_{N=1} - M(\gS_0)|_{N=1})$
of the initial value (we use $\mathrm{horizon\_frac} = 0.5$).

\subsection{Two complementary harnesses}
\label{sec:harnesses}

Because the ecosystem question has been operationalised in two very
different ways in the literature --- fine-tuning ecosystems
\citep{hodel2026epistemic, shumailov2024nature} and retrieval ecosystems
\citep{wang2025web} --- we run \emph{both} on the \emph{same} metric
suite, so that \mvp and axis conclusions can be cross-validated.

\subsubsection{E1: Fine-tuning ecosystem}
\label{sec:e1}

\para{Backbone.} \distilgpt (82M parameters), one fresh copy per agent.
\para{Data.} \wikitext-raw-v1 (train split as initial real data; test
split, fixed across generations, for perplexity).
\para{Per-ecosystem token budget.} 80{,}000 synthetic tokens per
generation, partitioned into $N$ equal slices, holding total compute
roughly constant across $N$.
\para{Loop.} For $t = 1, \ldots, T$:
(i) each agent generates $\mathrm{budget}/N$ tokens at temperature
$1.0$ and top-$p$ $0.95$; (ii) generated streams are pooled,
retokenised in 64-token blocks, shuffled, and re-split into $N$ new
equal slices; (iii) each agent fine-tunes one epoch on its slice
(AdamW, lr $5{\times}10^{-5}$, batch size $16$). Replace mode: no
real-data refresh.
\para{Sweep.} $N \in \{1, 2, 4, 8, 16\}$, $T = 6$, two seeds per
condition.

\subsubsection{E2: Retrieval-augmented (RAG) ecosystem}
\label{sec:e2}

\para{Models.} A pool of small instruct-tuned LLMs hosted on
OpenRouter: Llama-3.1-8B-Instruct, Mistral-7B-Instruct,
Qwen-2.5-7B-Instruct, Gemma-2-9B-it, GPT-4o-mini,
DeepSeek-Chat-v3, Hermes-3-Llama-3.1-8B, WizardLM-2-8x22B.
Larger $N$ exposes the ecosystem to more pretraining-prior
diversity.
\para{Internet seed.} 20 paragraphs of $\geq 200$ characters sampled
from \wikitext.
\para{Loop.} At each iteration each agent retrieves
$k = \lceil \beta \cdot |\mathrm{pool}| \rceil$ random posts
($\beta = 0.2$), uses them as RAG context, and emits one $\leq 120$-token
paragraph. New posts are appended to the pool.
\para{Sweep.} $N \in \{1, 2, 3, 5, 8\}$ (subset by API budget),
$T = 12$, three seeds per condition.

\subsubsection{E3: Diversity-axis comparison}
\label{sec:e3}

At fixed $N{=}4$ in the RAG harness, four conditions:
\begin{itemize}[leftmargin=*,itemsep=0pt,topsep=0pt]
    \item \axisseed --- same model, same prompt, agents differ only
    by sampling RNG (control).
    \item \axisdata --- same model, same prompt, each agent retrieves
    from a disjoint slice of the seed corpus.
    \item \axisprompt --- same base model, four distinct system-prompt
    personas (physicist, detective novelist, children's storyteller,
    tech journalist).
    \item \axisfamily --- four OpenRouter models from different
    pretraining lineages.
\end{itemize}
$T = 10$, three seeds per condition.

\subsection{Metrics}
\label{sec:metrics}

We evaluate every iteration of every run with five metrics
(\tabref{tab:metrics}). The semantic metrics use the
\minilm sentence encoder. The Hill--Shannon Diversity (\hsd) is the
Vendi score \citep{friedman2023vendi}, an effective number of distinct
outputs. ``Lower-better'' metrics (perplexity) and
``higher-better'' metrics (the four diversity scores) are handled
symmetrically by the \mvp definition above.

\begin{table}[t]
    \centering
    \caption{Evaluation metrics and their interpretation.}
    \label{tab:metrics}
    \small
    \begin{tabular}{@{}lll@{}}
        \toprule
        \textbf{Metric} & \textbf{Definition} & \textbf{Direction} \\
        \midrule
        perplexity        & $\exp(\mathrm{NLL})$ on \wikitext test       & lower better \\
        distinct-2        & unique bigrams / total bigrams               & higher better \\
        mean pairwise dist.\ & mean off-diag.\ of $1 - \cos(e_i, e_j)$  & higher better \\
        Frobenius (\citealt{wang2025web}) & $\|1 - \cos(e_i, e_j)\|_F$  & higher better \\
        \hsd (Vendi)      & $\exp(-\sum (\lambda/n) \log(\lambda/n))$    & higher better \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Statistical analysis}
\label{sec:stats}

For each (experiment, $N$, metric) we fit an ordinary-least-squares
linear regression of the metric on iteration index, report the slope
with $95\%$ confidence interval, and the two-sided $p$-value of the
slope's null. We define \mvp as in \secref{sec:problem}. All numbers
are mean $\pm$ standard deviation across seeds. Mean values aggregate
both across seeds and across the $N$ agents within each condition.

\subsection{Implementation and reproducibility}
\label{sec:repro}

Random seeds for Python, NumPy, PyTorch, and our slice-permutation RNG
are set explicitly per run. Per-iteration metrics and a sample of
generated text are streamed to JSONL logs for post-hoc inspection. Each
run's full configuration is checkpointed alongside its result.
Hardware: 1$\times$NVIDIA RTX A6000 (E1); CPU + OpenRouter API (E2,
E3). E2/E3 do not depend on bit-identical sampling; OpenRouter routes
models to different providers and the random seed controls only the
Python-level RNG (retrieval, prompt selection).
